`methyvim`: Targeted and Model-free Differential Methylation Analysis in

### 2018-06-21 paper/paper_BiocF1000.Rmd
update f1000; misc other edits

---
title: '`methyvim`: Targeted and Model-free Differential Methylation Analysis in
  R'
author:
  - name: Nima S. Hejazi
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Center for Computational Biology, University of California, Berkeley
    orcid: 0000-0002-7127-2789
    email: nhejazi@berkeley.edu
  - name: Rachael V. Phillips
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Department of Statistics, University of California, Berkeley
abstract: |
  We present `methyvim`, a general algorithm for the nonparametric estimation of
  treatment/exposure effects on DNA methylation at CpG sites, complete with
  straightforward statistical inference for such estimates. This approach
  leverages variable importance measures, a class of parameters that arise in
  causal inference and are defined in such a manner that they provide targeted
  estimates of the relative importance of individual CpG site with respect to
  a binary exposure/treatment assigned at the level of subjects. Additionally,
  this procedure is computationally efficient; incorporating a preliminary
  screening step to isolate a subset of sites for which there is cursory
  evidence of differential methylation and a multiple testing correction step
  to control the False Discovery Rate as if all sites were tested. This
  technique for analysis of differentially methylated positions provides an
  avenue to move beyond linear models without the loss of interpretable
  statistical inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, statistical variable importance, targeted minimum loss-based
  estimation, machine learning
bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina _Infinium_ Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without
restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e. confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging
[@libbrecht2015machine].

In this paper, we present an alternative such statistical analysis approaches in
the form of nonparametric estimation procedures that provide simple and well
interpretable statistical inference. Inspired by concepts of counterfactual
effects from the causal inference literature and machine learning algorithms, we
provide a computationally efficient technique for obtaining targeted estimates
of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set
of CpGs, controlling for the False Discovery Rate (FDR) as if all sites were
tested. Under counterfactual assumptions, these estimators have asymptotic
normal sampling distributions and thus provide reliable inference.
Counterfactuals are "counter"-fact, defined in terms of an unrealistic yet ideal
experiment. In the context of DNA methylation and for a binary treatment, we
define the counterfactual outcomes as the observed methylation value (whether
Beta- or M-values) a CpG site would have if all subjects were treated and the
methylation value a CpG site would have if all subjects were not treated. Of
course, these counterfactual outcomes are impossible to observe but they are
very helpful for us to asses the importance of a CpG site relative to a subjects
treatment/exposure and they can be statistically estimated
[@vdl2006statistical]. Our VIM definition uses the concepts of counterfactuals
and the framework for estimation is built on _targeted maximum likelihood
estimation_ or _targeted minimum loss-based estimation_ (__TMLE__)
[@vdl2006targeted]. This methodology targets the individual importance of each
CpG site by taking advantage of the prediction power of machine learning
algorithms. The mathematical properties of this approach have been explored in
[@tuglus2008targeted] and we focus on our contribution to it's application,
`methyvim`, an R package that make this technique for differential methylation
analysis easily accessible.

@@

@@

2. Nonparametric estimates of VIMs, for the specified target parameter, are
   computed at each of the CpG sites passing the screening step. The VIMs are
   defined in such a way that the estimated effects is of an exposure/treatment
   on the methylation status of a target CpG site, controlling for the observed
   methylation status of the neighbors of that site. Currently, routines are
   adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
   Future releases will support doubly-robust estimates of these VIMs (via the
   [`drtmle`](https://cran.r-project.org/package=drtmle) package)
   and add parameters for continuous treatments/exposures (via the
   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).

3. Since pre-screening is performed prior to estimating VIMs, we make use of a
   multiple testing correction uniquely suited to such settings. Due to the
   multiple testing nature of the estimation problem, a variant of the Benjamini
   & Hochberg procedure for controlling the False Discovery Rate (FDR) is
   applied [@benjamini1995controlling]. Specifically, we apply the "modified
   marginal Benjamini & Hochberg step-up False Discovery Rate controlling
   procedure for multi-stage analyses" (FDR-MSA), which is guaranteed to
   control the FDR as if all sites were tested (i.e., without screening)
   [@tuglus2009modified].

@@

For the CpG sites that passed the pre-screening step, the user chooses the
parameter of interest that will be computed at each site. _In all cases, an
estimator of the parameter of interest is constructed via targeted minimum
loss-based estimation_.

@@

Support for continuous-valued treatments or exposures is _planned but not yet
available_, though work is underway to incorporate into our methodology the
following

* A nonparametric variable importance measure (NPVI) [@chambaz2012estimation]:
  The effect of continuous-valued exposure or treatment (the observed
  methylation at a target CpG site) on an outcome of interest is estimated,
  controlling for the observed methylation at all other CpG sites in the same
  neighborhood as the target (treatment) site, based on a parameter that
  compares values of the treatment against a reference value taken to be the
  null. In particular, the implementation provided is designed to assess the
  effect of differential methylation at the target CpG site on a (typically)
  phenotype-level outcome of interest (e.g., survival), in effect providing an
  nonparametric evaluation of the impact of methylation at the target site on
  said outcome.

@@

```{r setup-minfidata}
suppressMessages(library(minfiData))
data(MsetEx)
mset <- mapToGenome(MsetEx)
grs <- ratioConvert(mset)
grs
```
We can create an object of class `methytmle` from any `GenomicRatioSet` object
simply invoking the S4 class constructor:

```{r make-methytmle}
library(methyvim)
mtmle <- .methytmle(grs)
```
Additionally, a `GenomicRatioSet` can be created from a matrix with the
function `makeGenomicRatioSetFromMatrix` provided by the `minfi` package.

@@

```{r, minfidata-methyvim}
suppressMessages(
  methyvim_cancer_ate <- methyvim(data_grs = grs, var_int = var_int,
                                  vim = "ate", type = "Beta", filter = "limma",
                                  filter_cutoff = 0.20, obs_per_covar = 2,
                                  parallel = FALSE, sites_comp = 125,
                                  tmle_type = "glm"
                                 )
)
```
Note that we set the `obs_per_covar` argument to a relatively low value (2,
where the recommended default is 20) for the purposes of this example as the
sample size is only 10. We do this only to exemplify the estimation procedure
and it is important to point out that such low values for `obs_per_covar` will
compromise the quality of inference obtained because this setting directly
affects the definition of the target parameter.

@@

We can view a table of results by examining the `vim` slot of the produced
`methytmle` object.

```{r vim-cancer-ate}
head(slot(methyvim_cancer_ate, "vim"))
```

@@

Here we introduce the R package `methyvim`, an algorithm for differential
methylation analysis that moves beyond linear models. Additionally, this
estimation procedure produces straightforward statistical inference. Inspired by
concepts of counterfactual effects from the causal inference literature and
machine learning algorithms, we provide a computationally efficient technique
for obtaining targeted estimates of nonparametric variable importance measures
for a pre-screened set of CpGs, controlling for the False Discovery Rate as if
all sites were tested, and the estimation framework is built on targeted
minimum loss-based estimation. By taking advantage of the prediction power of
machine learning algorithms, this methodology targets the individual importance
of each CpG site, providing a fully nonparametric statistical model for the
analysis of differential methylation.

@@

Archived source code as at time of
publication: https://github.com/nhejazi/methyvim

>>>>

---
title: '`methyvim`: Targeted and Model-free Differential Methylation Analysis in
  R'
author:
  - name: Nima S. Hejazi
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Center for Computational Biology, University of California, Berkeley
    orcid: 0000-0002-7127-2789
    email: nhejazi@berkeley.edu
  - name: Rachael V. Phillips
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Department of Statistics, University of California, Berkeley
    orcid: 0000-0002-XXXX-XXXX 
abstract: |
  We present `methyvim`, an R package implementing a general algorithm for the
  nonparametric estimation of treatment effects on DNA methylation at CpG sites
  throughout the genome, complete with straightforward statistical inference for
  such estimates. The approach leverages variable importance measures derived
  from statistical parameters arising causal inference, defined in such a manner
  that they may be used to obtain targeted estimates of the relative importance
  of individual CpG site with respect to a binary treatment assigned at the
  phenotype level. The procedure implemented is computationally efficient,
  incorporating a preliminary screening step to isolate a subset of sites for
  which there is cursory evidence of differential methylation as well as a
  unique multiple testing correction to control the False Discovery Rate with
  the same rigor as if all sites were tested. This technique for analysis of
  differentially methylated positions provides an avenue to incorporate flexible
  state-of-the-art machine learning algorithms into the estimation of
  differential methylation effects without the loss of interpretable statistical
  inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, statistical variable importance, targeted minimum loss-based
  estimation, machine learning
bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA methylation occurs at CpG sites
and involves the addition of a methyl group ($\text{CH}_3$) to the fifth carbon
of the cytosine ring structure to form 5-methylcytosine. Numerous biological and
medical studies have implicated DNA methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to rigorously probe the molecular mechanisms of this
epigenetic process. Modern assays, like the Illumina _Infinium_ Methylation chip
technology, allow for quantitative interrogation of DNA methylation at a large
set of CpG sites scattered across the genome at single-nucleotide resolution;
moreover, much effort has been invested, by the bioinformatics community, in the
development of tools for properly removing technological effects that may
contaminate biological signatures measured by such assays. Despite these
advances in both biological and bioninformatical techniques, most statistical
methods available for differential analysis of data produced by such assays rely
on over-simplified models that fail to incorporate many of the recent advances
in machine learning. Generally, typical statistical estimation approaches do not
readily extend to such high-dimensional data without restrictive modeling
assumptions and the use of inferentially costly hypothesis testing corrections.
When these standard assumptions are violated, estimates of the population-level
effect of an exposure or treatment may suffer from large bias. Data-adaptive
estimation procedures that utilize machine learning can control for confounding
even in high-dimensional settings; however, interpretable statistical inference
(i.e., confidence intervals and hypothesis tests) resulting from these
data-adaptive estimates is challenging to obtain [@libbrecht2015machine].

In this paper, we briefly present an alternative to such statistical analysis
approaches in the form of a nonparametric estimation procedure that provides
simple and readily interpretable statistical inference, discussing at length a
recent implementation of the methodology in the `methyvim` R package. Inspired
by recent advances in statistical causal inference and machine learning, we
provide a computationally efficient technique for obtaining targeted estimates
of nonparametric _variable importance measures_ (__VIMs__), estimated at a set
of pre-screened CpG sites, controlling for the False Discovery Rate (FDR) as if
all sites were tested. Under standard assumptions (e.g., identifiability, strong
ignorability) [@pearl2009causality], targeted minimum loss-based estimators of
regular asymptotically linear estimators have sampling distributions that are
asymptotically normal, allowing for reliable point estimation and the
construction of Wald-style confidence intervals [@vdl2011targeted,
@vdl2018targeted]. As counterfactuals are contrary-to-fact -- that is, defined
in terms of missing data -- in the context of DNA methylation, we define the
counterfactual outcomes under a binary treatment as the observed methylation
(whether Beta- or M-) value a CpG site would have if all subjects were given the
treatment and the methylation value a CpG site would have if treatment were
withheld from all subjects. Although these counterfactual outcomes are, of
course, impossible to observe, they have statistical analogs that may be
reliably estimated from observed data [@pearl2009causality]. We describe an
algorithm that incorporates the estimation of VIMs using _targeted minimum
loss-based estimation_ (__TMLE__) [@vdl2006targeted], deferring detailed
description and analysis of the statistical methodology to work-in-progress
[@hejazi2018methylation]. This methodology assesses the individual importance of
a given CpG site by utilizing state-of-the-art machine learning algorithms in
the estimation of a targeted VIM, initially proposed and explored in
[@tuglus2008targeted]. In the presnet work, we focus on a software package,
`methyvim`, that implements a variant of this methodology, specifically tailored
to differential methylation analysis.

@@

## Data Structure

We consider an observed data structure $O = (W_k, A, Y_j)$, where $Y_j, j = 1,
\ldots J$ represents a given CpG site of interest, $A \in \{0, 1\}$ represents a
binary phenotype-level treatment, and $W_k, k = 1, \ldots K(j)$ is a matrix of
the observed values of CpG sites in the same neighborhood as $j$. We consider
the case of observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where we
take $O \sim \mathcal{P} \in \mathcal{M}$, which is to say that $O$ is governed
by an unknown probability distribution $\mathcal{P}$, taken only to reside in a
nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
data-generating process. We consider having access to measurements on a large
number $G$ of CpG sites (e.g., 850,000, as measured by the Illumina
MethylationEPIC BeadChip arrays), and as a matter of practicality, let $J \leq
G$, so that $J$ indexes only those CpG sites that pass an arbitrary
pre-screening procedure; note that the equality is only obtained in the case
that all sites pass the pre-screening procedure. Further, we let $K(j)$ index
the set of all CpG sites so that $k$, the realization of $K(j)$ for a given CpG
site $j$, is a set of the CpG sites that neighbor the CpG site $j$, where the
definition of a neighborhood is left as user-specified; thus, $W_k$ is a matrix
of observed methylation values for the set of CpG sites $k$ that neighbor a
particular target site $j$. We consider estimating a target parameter ($\psi_j$)
as a variable importance measure of the effect of changing a treatment $A$ on
the methylation of a CpG site $j$, accounting for the observed methylation at
the set of sites $k$ that neighbor the site $j$. After a complete run of the
procedure, we have access to $j$ VIM estimates $\psi_j$, each corresponding to
the degree to which a given CpG site $j$ was responsive to a change in the
binary treatment $A$, after controlling for any confounding exerted by its
neighboring CpG sites. When machine learning estimators are employed in
estimating $\psi_j$, this produces a nonparametric variable importance measure
of differential methylation, allowing for the identification of differentially
methylated positions (DMPs).

@@

2. Nonparametric estimates of VIMs, for the specified target parameter, are
   computed at each of the CpG sites passing the screening step. The VIMs are
   defined in such a way that the estimated effects is of an exposure/treatment
   on the methylation status of a target CpG site, controlling for the observed
   methylation status of the neighbors of that site. Currently, routines are
   adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
   Future releases will support doubly-robust estimates of these VIMs (via the
   [`drtmle`](https://cran.r-project.org/package=drtmle) package)
   and add parameters for continuous treatments/exposures (via the
   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) and
   [`txshift`](https://github.com/nhejazi/txshift) R packages).

3. Since pre-screening is performed prior to estimating VIMs, we make use of a
   multiple testing correction uniquely suited to such settings. Due to the
   multiple testing nature of the estimation problem, a variant of the Benjamini
   and Hochberg procedure for controlling the False Discovery Rate (FDR) is
   applied [@benjamini1995controlling]. Specifically, we apply a modified
   marginal Benjamini and Hochberg step-up False Discovery Rate controlling
   procedure for multi-stage analyses (FDR-MSA), which has established
   theoretical guarantees to control the FDR at the same rate as if all sites
   were tested (i.e., without screening) [@tuglus2009modified].

@@

For the CpG sites that pass the pre-screening step, a user-specified target
parameter of interest is estimated independently at each site. _In all cases,
an estimator of the parameter of interest is constructed via targeted minimum
loss-based estimation_.

@@

Support for continuous-valued treatments or exposures is _planned but not yet
available_. Future releases will allow users to assess continuous-valued
treatments by relying on parameters estimable through implementations available
in the following software packages:

* `tmle.npvi` - A nonparametric variable importance measure (NPVI)
  [@chambaz2012estimation]. The effect of a continuous-valued exposure or
  treatment (the observed methylation at a target CpG site) on an outcome of
  interest is estimated, controlling for the observed methylation value at all
  other CpG sites in the same neighborhood as the target site. This uses a
  parameter that compares values of the treatment against a user-specified
  reference value taken to be the null. In particular, the implementation to be
  provided is designed to assess the effect of differential methylation at the
  target CpG site on an outcome of interest (e.g., survival), providing a
  nonparametric evaluation of the impact of methylation at the target site.

* `txshift` - Variable importance using a causal parameter defined as the
  counterfactual outcome under a posited shift of a continuous-value treatment
  of interest using stochastic intervention policies [@munoz2012population,
  diaz2018stochastic, hejazi2018txshift]. The value of an outcome of interest
  under an unobserved value of the (continuous-valued) treatment, specified
  through a user-provided _additive shift_ of the observed treatment, may be
  data-adaptively estimated. This allows for the effect of changes/shifts in the
  observed methylation at a CpG site on an outcome of interest to be evalauted,
  providing a nonparametric evalution of the relative importance of a particular
  target CpG site with respect to another variable measured in the same study.

@@

```{r setup-minfidata}
suppressMessages(library(minfiData))
data(MsetEx)
mset <- mapToGenome(MsetEx)
grs <- ratioConvert(mset)
grs
```
We can create an object of class `methytmle` from any `GenomicRatioSet` object
simply invoking the S4 class constructor `.methytmle`:

```{r make-methytmle}
library(methyvim)
grs_mtmle <- .methytmle(grs)
grs_mtmle
```
Additionally, a `GenomicRatioSet` can be created from a matrix with the
function `makeGenomicRatioSetFromMatrix` provided by the `minfi` package.

@@

```{r, minfidata-methyvim, warning=FALSE}
methyvim_cancer_ate <- methyvim(data_grs = grs, var_int = var_int,
                                vim = "ate", type = "Beta", filter = "limma",
                                filter_cutoff = 0.20, obs_per_covar = 2,
                                parallel = FALSE, sites_comp = 100,
                                tmle_type = "glm"
                               )
```
Note that we set the `obs_per_covar` argument to a relatively low value (2,
where the recommended default is 20) for the purposes of this example as the
sample size is only 10. We do this only to exemplify the estimation procedure
and it is important to point out that such low values for `obs_per_covar` will
compromise the quality of inference obtained because this setting directly
affects the definition of the target parameter.

@@

We can view a table of results by examining the `vim` slot of the produced
object, displayed using the accessor method of the same name:

```{r vim-cancer-ate}
vim(methyvim_cancer_ate)
```

Note that simply printing the object will also display an abbreviated version of
these same results, alongside a bit of additional information and all of the
standard information normally printed for the `GenomicRatioSet` superclass.

@@

Here we introduce the R package `methyvim`, an implementation of a general
algorithm for differential methylation analysis that allows for recent advances
in causal inference and machine learning to be leveraged in computational
biology settings. The estimation procedure produces straightforward statistical
inference and takes great care to ensure computationally efficiency of the
technique for obtaining targeted estimates of nonparametric variable importance
measures. The software package includes techniques for pre-screening a set of
CpG sites, controlling for the False Discovery Rate as if all sites were tested,
and for visualzing the results of the analyses in a variety of ways. The anatomy
of the software package is dissected and the design described in detail. The
`methyvim` R package is available via the Bioconductor project.

@@

Archived source code as at time of publication:
https://github.com/nhejazi/methyvim/releases/tag/v1.3.1
